---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE
)
```

# tsrecipes

<!-- badges: start -->
<!-- badges: end -->

The goal of tsrecipes is to provide time series proprocessing
to accommodate time series classification and clustering in the
[tidymodels](https://www.tidymodels.org/) framework.

The primary steps are 

* discrete cosine transform for dimensionality reduction and feature engineering
* dynamic time warping clustering

## Installation

You can install tsrecipes with:

```r
# install.packages("devtools")
devtools::install_github("tmastny/tsrecipes")
```

## Examples

### Time series dimensionality reduction

In time series classification, using the time series as a feature
often results in poor accuracy because
it's the auto-correlation between entries and the overall trend
that's the most informative, rather than individual values.

The discrete cosine transform is one way to extract useful, 
uncorrelated features with significantly fewer dimensions than the time series.

In this example, it's possible to classify time series of length 1751 into 
4 classes with 70% accuracy, only using 16 dimensions.

```{r}
library(tidymodels)
library(tsrecipes)

rec <- recipe(ethanol, var = names(ethanol), roles = c("id", "outcome", "input")) %>%
  step_dct(ts, k = tune())

set.seed(2532)
tune_results <- workflow() %>%
  add_model(multinom_reg() %>% set_engine("nnet")) %>%
  add_recipe(rec) %>%
  tune_grid(
    resamples = validation_split(ethanol),
    grid = expand_grid(k = c(4, 8, 16, 32, 64))
  )

tune_results %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  select(k, mean)
```

### Time series clustering

tsrecipes also enables times series clustering using the
dynamic time warping metric. `step_dtw` is powered by the
[dtwclust](https://github.com/asardaes/dtwclust) package.

```{r, echo=FALSE}
ethanol_clusters <- readRDS("vignettes/ethanol_clusters.RDS")
```

```{r, eval=FALSE}
ethanol_clusters <- recipe(ethanol) %>%
  step_dtw(ts, k = 4) %>%
  prep() %>%
  bake(ethanol)
```

```{r}
ethanol_clusters %>%
  mutate(n = list(1:1751)) %>%
  unnest(c(ts, n)) %>%
  ggplot() +
  geom_line(aes(n, ts, group = id), alpha = 0.2, show.legend = FALSE) +
  facet_wrap(~dtwclust_ts)
```

## References

Sayood, K. (2006). *Introduction to data compression*.

Primer on DCT: https://squidarth.com/rc/math/2018/06/24/fourier.html

dtwclust 
and 
dtt


This package is modeled after the [textrecipes](https://github.com/tidymodels/textrecipes):
both packages transform a sequence of data (words vs. time series).

The [feasts](https://feasts.tidyverts.org/) R package and the
Python package [tsfresh](https://github.com/blue-yonder/tsfresh)
contain many other time series preprocessing methods not included in this package.

